Containers and R: the Rockerverse and beyond

### 2019-08-27 manuscript.Rmd
Update manuscript.Rmd
Added info about my affiliation

---
title: "Containers and R: the Rockerverse and beyond"
author:
  - name: Daniel Nüst
    affiliation: Institute for Geoinformatics, University of Münster
    address:
    - Heisenbergstr. 2, 48149 Münster, Germany
    email: daniel.nuest@uni-muenster.de
  - name: Carl Boettiger
    #affiliation: -
    #email: -
  - name: Robrecht Cannoodt
    #affiliatin: -
    #email: -
  - name: Dirk Eddelbuettel
    #affiliation:
    #email: -
  - name: Mark Edmondson
    affiliation: IIH Nordic A/S, Google Developer Expert for GCP
    email: mark@markedmondson.me
  - name: Colin Fay
    #affiliation: -
    #email: -
  - name: Ben Marwick
    #affiliation: -
    #email: -
  - name: Karthik Ram
    #affiliation: -
    #email: -
  - name: Noam Ross
    #affiliation: -
    #email: -
  - name: Nan Xiao
    #affiliation: -
    #email: -
  - name: Lori Shepherd
    affiliation: Roswell Park Comprehensive Cancer Center
    address:
    - Elm & Carlton Streets, Buffalo, New York, 14263 USA
    email: lori.shepherd@roswellpark.org
  - name: Nitesh Turaga
    affiliation: Roswell Park Comprehensive Cancer Center
    address:
    - Elm & Carlton Streets, Buffalo, New York, 14263 USA
    email: nitesh.turaga@roswellpark.org
abstract: >
  The Rocker project provides widely-used Docker images for R across different application scenarios.
  This articles surveys describe downstream projects building upon Rocker.
  We also look beyond Rocker to other projects connecting containerisation with R.
  These use cases and the diversity of applications demonstrate the power of Rocker and containerisation for collaboration, effectivity, scalability, and transparency.
preamble: >
   \usepackage{longtable}
output: rticles::rjournal_article
---

@@

- [MRO](https://github.com/nuest/mro-docker)
- [Renjin](https://github.com/nuest/renjin-docker)
- [pqR](https://github.com/nuest/pqr-docker)
- [FastR](https://github.com/nuest/fastr-docker)

- [Bioconductor](http://bioconductor.org/help/docker/), https://github.com/Bioconductor/bioc_docker

#
# Rough draft - they asked one sentance per line
#

_Bioconductor_ is an open source, open development project for the analysis and comprehension of genomic data [1].
The project consists of 1741 R software packages as of August 15th 2019, as well as packages containing annotation or experiment data.
_Bioconductor_has a semiannual release cycle, each release is associated with a particular version of R.
Docker images allow availability of current and past versions of _Bioconductor_ for convenience and reproducibility.

_Bioconductor_ 'base' docker images are built on top of rocker/r-ver and rocker/rstudio.
_Bioconductor_ installs packages based on the R version, and therefore uses rocker/rstudio and rocker/r-ver version tagging.
_Bioconductor_ selects the desired version of R from rocker, adds the BiocManager CRAN package for installing appropriate versions of _Bioconductor_ packages, and creates a _Bioconductor_ docker image with an informative tag (R\_version\_Bioc\_version).
The images are summarized on the _Bioconductor_ web site [5], maintained in github [6], and available to the community through DockerHub [3].
Past and current combinations of R and _Bioconductor_ are therefore accessible via a specific docker tag.

_Bioconductor_ has several images in addition to 'base', specific to various areas of research.
The 'core' image installs the most commonly used _Bioconductor_ packages.
_Bioconductor_ images for proteomics, metabolomics, and flow cytometry are community maintained.
All community maintained images build on top of the _Bioconductor_ base image and therefore indirectly the rocker images [4].
To simplify building and maintaining _Bioconductor_ images, we use a Ruby templating engine.

A recent audit of our Dockerfiles, following best practices from the Docker website, led to a reduction in the size and number of layers in images we produce.
The most important insights involve the 'union' file system used by Docker [REF:FIXME].
In this file system, once a layer (e.g., `RUN` statement) writes to a file path, the file path is never altered.
A subsequent layer that might appear to remove or overwrite the path actually masks, rather than alters, the original.
It is therefore important to clean up (e.g., cache removal) within each layer, and to avoid re-installing existing dependencies.

A recent innovation is to produce a `bioconductor_full` image to emulate the _Bioconductor_ nightly Linux build machine.
The image contains the _system dependencies_ needed to install and check almost all (1730 of 1741) _Bioconductor_ software packages.
Users no longer have to manage complciated system dependencies.
The image is configured so that `.libPaths()` has `/usr/local/lib/R/host-site-library` as the first location.
Users mounting a location on the host file system to this location then persist installed packages across docker sessions or updates.
Many _R_ users pursue flexible work flows tailored to particular analysis needs, rather than standardized work flows.
The `bioconductor_full` image is well-suited to this pattern.
`bioconductor_full` provides developers with a test environment like _Bioconductor_'s build system.

Use of images suggests several interesting possibilities for the _Bioconductor_ project.
Images may be valuable in teaching, where participants pull pre-built images to avoid complicated configuration of their own computing environemnts.
An appeal of this over our current approach (providing Amazon Machine Instances for the duration of the course) is the utility of the image to participant after the course is over.
`bioconductor_full` introduces a common system configuration, so it becomes increasingly sensible for _Bioconductor_ to distribute convenient _binary_ packages.
Images also suggest approaches to more advanced computational models.
For instance, we are exploring use of images for Helm [REF:FIXME] orchestrated Kubernetes [REF:FIXME] clusters on the Google Cloud Platform.
The user interacts with a manager image based on `bioconductor_full`, configured to perform map-reduce style computations via the BiocParallel package communicating with minimally-configured worker images.
A strength of this approach is that the responsibility for complex software configuration (including customized development) is shifted from the user to the experienced _Bioconductor_ core team.

#
# Need to move these to bib file and properly format.
#

[1]  Gentleman, Robert C.; Carey, Vincent J.; Bates, Douglas M.; Bolstad, Ben;
Dettling, Marcel; Dudoit, Sandrine; Ellis, Byron; Gautier, Laurent; Ge,
Yongchao; Gentry, Jeff; Hornik, Kurt; Hothorn, Torsten; Huber, Wolfgang; Iacus,
Stefano; Irizarry, Rafael; Leisch, Friedrich; Li, Cheng; Maechler, Martin;
Rossini, Anthony J.; Sawitzki, Gunther; Smith, Colin; Smyth, Gordon; Tierney,
Luke; Yang, Jean Y. H.; Zhang, Jianhua (2004). "_Bioconductor_: open software
development for computational biology and bioinformatics". Genome Biology. 5
(10): R80. doi:10.1186/gb-2004-5-10-r80. PMC 545600

[2] R Core Team (2019). R: A language and environment for statistical
  computing. R Foundation for Statistical Computing, Vienna, Austria.
  URL https://www.R-project.org/

[3] ??DockerHub citation?

[4] ??Reference our docker help page/github?

[5] ?? https://bioconductor.org/help/docker/

[6] ?? https://github.com/Bioconductor/bioc_docker

[*] The original work on `bioc_docker` was done by Dan Tanenbaum

>>>>

---
title: "Containers and R: the Rockerverse and beyond"
author:
  - name: Daniel Nüst
    affiliation: Institute for Geoinformatics, University of Münster 
    address:
    - Heisenbergstr. 2, 48149 Münster, Germany
    email: daniel.nuest@uni-muenster.de
  - name: Carl Boettiger
    #affiliation: -
    #email: -
  - name: Robrecht Cannoodt
    #affiliatin: -
    #email: -
  - name: Dirk Eddelbuettel
    #affiliation:
    #email: -
  - name: Mark Edmondson
    affiliation: IIH Nordic A/S, Google Developer Expert for GCP
    email: mark@markedmondson.me
  - name: Colin Fay
    affiliation: ThinkR
    email: contact@colinfay.me
  - name: Ben Marwick
    #affiliation: -
    #email: -
  - name: Karthik Ram
    #affiliation: -
    #email: -
  - name: Noam Ross
    #affiliation: -
    #email: -
  - name: Nan Xiao
    #affiliation: -
    #email: - 
abstract: >
  The Rocker project provides widely-used Docker images for R across different application scenarios.
  This articles surveys describe downstream projects building upon Rocker.
  We also look beyond Rocker to other projects connecting containerisation with R.
  These use cases and the diversity of applications demonstrate the power of Rocker and containerisation for collaboration, effectivity, scalability, and transparency.
preamble: >
   \usepackage{longtable}
output: rticles::rjournal_article
---

@@

- [MRO](https://github.com/nuest/mro-docker)
- [Renjin](https://github.com/nuest/renjin-docker)
- [pqR](https://github.com/nuest/pqr-docker)
- [FastR](https://github.com/nuest/fastr-docker)
- [Bioconductor](http://bioconductor.org/help/docker/), https://github.com/Bioconductor/bioc_docker
